---
title: "Exercises"
linktitle: "6. Analyze your data"
date: "2020-08-17"
publishdate: "2020-08-17"
menu:
  exercise:
type: docs
output:
  blogdown::html_page:
    toc: true
    toc_depth: 2
    number_sections: true
    pandoc_args: ["--number-offset=5"]
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  error = TRUE,
  collapse = TRUE,
  comment = "#> ",
  class.output = "output",
  class.message = "message",
  class.error = "message"
)
```
# Analyze your data


## Challenge

I've given instructions for a project using the NLSY data. If you have data that you're working with and you'd like to practice with that data instead, go for it! Instead of the variables mentioned below, use whatever variables from your dataset you want. For example, change categorical variables that are currently numeric to factors as needed in your data. Then restrict your sample based on whatever conditions make sense to you for a project you're working on (or want to work on), and run a regression with whichever variables interest you!

### Prepare your project {-}

- File -> New Project -> New Directory -> New Project
- Name it something like NLSY and put it in an appropriate folder on your computer
- Within that folder, make new folders as follows:

```
NLSY/
 ├─ NLSY.Rproj
 ├─ data/
 │   ├── raw/
 │   └── processed/
 ├─ code/
 └── results/
     ├── tables/
     └── figures/
```

### Prepare the data {-}

- Copy and paste `nlsy.csv` into `data/raw`.
- Create a new file and save it as `clean_data.R`.
- In that file, read in the NLSY data and load any packages you need. Make sure you replace any missing values with `NA`. Hint: there are extra missing values in the `age_bir` variable. Also, the variable names might be useful: 

```{r}
colnames_nlsy <- c(
  "glasses", "eyesight", "sleep_wkdy", "sleep_wknd",
  "id", "nsibs", "samp", "race_eth", "sex", "region",
  "income", "res_1980", "res_2002", "age_bir"
)
```
  - Add factor labels to `eyesight`, `sex`, `race_eth`, `region`, as in earlier slides. Select those variables plus `income`, `id`, `nsibs`, `age_bir`, and the sleep variables. Then restrict to complete cases and people with incomes < $30,000. Make a variable for the log of income (replace with `NA` if income <= 0).
  - Also in that file, save your new dataset as a `.rds` file to the `data/processed` folder.
  
### Do some exploratory analysis {-}

- Create a file called `create_figure.R`. In this file, read in the cleaned dataset. Load any packages you need. Then make a `ggplot` figure of your choosing to show something about the distribution of the data. Save it to the `results/figures` folder as a `.png` file using the `ggsave()` function.
- Create a file called `table_1.R`. In this file, read in the cleaned dataset and use the `tableone` package to create a table 1 with the variables of your choosing. Modify the following code to save it as a `.csv` file. Open it in Excel/Numbers/Google Sheets/etc. to make sure it worked.

```{r, eval = FALSE}
tab1 <- CreateTableOne(...) %>% print() %>% as_tibble(rownames = "id")
write_csv(tab1, ...)
```

### Do some regression analysis {-}

- In another file called `lin_reg.R`, read in the data and run the following linear regression: `lm(log_inc ~ age_bir + sex + race_eth + nsibs, data = nlsy)`. Modify the CI function to produce a table of results for a *linear* regression. Add an argument `digits = `, with a default of 2, to allow you to choose the number of digits you'd like. Save it in a separate file called `functions.R`. Use `source()` to read in the function at the beginning of your script.
- Save a table of your results as a `.csv` file. Make the names of the coefficients nice!
- Using the results, use `ggplot` to make a figure. Use `geom_point()` for the point estimates and `geom_errorbar()` for the confidence intervals. It will look something like this:

```{r, eval = FALSE}
ggplot(data) + 
  geom_point(aes(x = , y = )) + 
  geom_errorbar(aes(x = , ymin = , ymax = ))
```

- Save that figure as a `.pdf` using `ggsave()`. You may want to play around with the `height = ` and `width = ` arguments to make it look like you want.