---
title: "Week 5 Exercises"
date: "2020-08-10"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: vignette
    toc: true
    toc_depth: 2
    number_sections: true
    pandoc_args: ["--number-offset=4"]
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  error = TRUE,
  collapse = FALSE,
  comment = "#> ",
  class.output = "output",
  class.message = "message"
)
```
# Functions


## Write a function to calculate exponents

### Square a number {-}

You're tired of writing `x^2` when you want to square `x`, so you want a function to square a number. You can call it `square()`. I showed this in the slides, now try on your own!

```{r, eval = FALSE}
# start out with a number to test
x <- 3
# you'll want your function to return this number
x^2
square <- function() {
  
}
# test it out
square(x)
square(53)
53^2 # does this match?
```

### Raise to any power {-}

You don't just want to square numbers, you want to raise them to higher powers too. Make a function that uses two arguments, `x` for a number, and `power` for the power. Call it `raise()`.

```{r, eval = FALSE}
raise <- function() {
  
}

# test with
raise(x = 2, power = 4)
# should give you
2^4
```

### Default arguments {-}

Change your `raise()` function to default to squaring `x` when the user doesn't enter a value for `power`.

```{r, eval = FALSE}

# test
raise(x = 5)
# should give you
5^2
```

## Change the grouping variable when summarizing

Notice that we didn't need any packages or data to write the functions above. Now we'll need our usuals, though:

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
nlsy <- read_csv("nlsy_cc.csv")
```

### Grouping argument {-}

We want to run code like this for multiple variables:
```{r}
nlsy %>% group_by(sex) %>% 
  summarize(mean_inc = mean(income), .groups = "drop")
nlsy %>% group_by(glasses) %>% 
  summarize(mean_inc = mean(income), .groups = "drop")
```

Write a function to calculate the stratified mean income for grouping variable `var`. In other words, write a function such that `mean_group_inc(var = "sex")` and `mean_group_inc(var = "glasses")` produce the results above.

Look at the function from the slides for help:
```{r}
var_q <- function(q, var) {
  quant <- nlsy %>%
    rename(new_var = var) %>% #<<
    summarize(q_var = quantile(new_var, probs = q))
  return(quant)
}
var_q(q = 0.5, var = "income")
```

Write your function here:
```{r}
mean_group_inc <- function(var) {
  
}

# test with
mean_group_inc(var = "glasses")
mean_group_inc(var = "sex")
```

### Grouping and summarizing arguments {-}

Rewrite your function to accept two arguments: `group_var` to determine what the grouping variable is, and `mean_var` to determine what variable you want to take the mean of (e.g., `mean_group(group_var = "sex", mean_var = "income")` should give you the same results as above).

```{r}
mean_group <- function(group_var, mean_var) {
  
}

# test with
mean_group(group_var = "sex", mean_var = "income")
```

## For loops

### Write a for loop {-}

We used this function:
```{r}
var_q_new <- function(q, var) {
  quant <- nlsy %>%
    rename(new_var = var) %>%
    summarize(q_var = quantile(new_var, probs = q)) %>%
    pull(q_var)
  return(quant)
}
var_q_new(q = 0.5, var = "income")
```
inside of a for loop in order to calculate each decile of income:
```{r}
qs <- seq(0.1, 0.9, by = 0.1)
deciles <- rep(NA, length(qs))
for (i in seq_along(qs)) {
  deciles[[i]] <- var_q_new(q = qs[[i]], 
                        var = "income")
}
deciles
```
Try changing `seq_along(qs)` to `1:5` and run the whole chunk again. What does `deciles` print now?

Change the for loop above to loop over different variables instead of different quantiles. That is, calculate the 0.25 quantile for each of `c("income", "age_bir", "nsibs")` in a for loop.

```{r, eval = FALSE}
vars <- c("income", "age_bir", "nsibs")
q_25s <- ...


```


### Nested loops {-}

You can nest for loops inside each other, as long as you use different iteration variables. Write a nested for loop to iterate over variables (with `i`) and quantiles (with `j`). You'll need to start with an empty matrix instead of a vector, with rows indexed by `i` and columns by `j`. Calculate each of the deciles for each of the above variables.

```{r, eval = FALSE}
vars <- ...
qs <- ...
results_mat <- matrix(NA, ncol = ..., nrow = ...)


```

