---
title: "Week 4 Exercises"
date: "2020-08-03"
output:
  html_document:
  # prettydoc::html_pretty:
  #   theme: architect
  #   highlight: vignette
    toc: true
    toc_depth: 2
    number_sections: true
    pandoc_args: ["--number-offset=3"]
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  error = TRUE,
  collapse = TRUE,
  comment = "#> ",
  class.output = "output",
  class.message = "message"
)
```
# Grouping and tables

Yup, still the same old package and data, though we'll be introducing another package today!
```{r, message = FALSE, warning = FALSE}
library(tidyverse)
nlsy <- read_csv("nlsy_cc.csv")
```

## Using pipes

Compare these two ways of making the same dataset:
```{r}
nlsy2 <- mutate(nlsy, 
                only = case_when(
                  nsibs == 0 ~ "yes",
                  TRUE ~ "no")
                )
nlsy3 <- select(nlsy2, 
                id, contains("sleep"), only)
only_kids <- filter(nlsy3, 
                    only == "yes")
```

vs:

```{r}
only_kids <- filter(
  select(
    mutate(nlsy,
      only = case_when(
        nsibs == 0 ~ "yes",
        TRUE ~ "no"
      )
    ),
    id, contains("sleep"), only
  ),
  only == "yes"
)
```

vs: 
```{r}
only_kids <- nlsy %>%
  mutate(only = case_when(
                nsibs == 0 ~ "yes",
                TRUE ~ "no"
                )
         ) %>%
  select(id, contains("sleep"), only) %>%
  filter(only == "yes")
```

All 3 techniques -- making new datasets, nesting functions, and using pipes -- can be useful, and which one is easier to write and read depends on the situation. Today we'll practice using pipes.

### Translate code {-}

Consider the following instructions for making a dataset:

1. Create a variable called `slp_cat_wkday` categorizing weekday sleep.
2. Subset the dataset to only include people with missing values.
2. Remove non-sleep-related variables from the dataset.
4. Call your finished dataset `missing_sleep`.

I've followed these instructions here.
```{r}
nlsy2 <- mutate(nlsy, slp_cat_wkdy = case_when(
  sleep_wkdy < 5 ~ "little",
  sleep_wkdy < 7 ~ "some",
  sleep_wkdy < 9 ~ "ideal",
  sleep_wkdy < 12 ~ "lots",
  TRUE ~ NA_character_))
missing_sleep <- filter(nlsy2, is.na(slp_cat_wkdy))
missing_sleep <- select(missing_sleep, starts_with("slp"), contains("sleep"))
missing_sleep
```

Now follow the instructions using pipes.

```{r, eval = FALSE}
missing_sleep <- nlsy %>% ...
```

### Order of operations {-}

Experiment with switching up the "order of operations". Can you complete the instructions in a different order and get the same result? Can you think of a situation when you might not be able to do so?

```{r}

```

## Summary statistics

We can use the `summarize()` function to calculate summary statistics:
```{r}
summarize(nlsy, 
          med_age_bir = median(age_bir),
          cor_sleep = cor(sleep_wkdy, sleep_wknd),
          ten_pctle_inc = quantile(income, probs = 0.1),
          ninety_pctle_inc = quantile(income, probs = 0.9))
```

It always helps to read the documentation when you learn a new function! Run this in your RStudio session.
```{r, eval = FALSE}
help(summarize)
```

Note from the documentation that you can also spell the function `summarise()`. Even though I consistently use American spellings in every other aspect of my life, you will find me spelling it like this often!

```{r}
nlsy %>% summarise(q.1 = quantile(age_bir, probs = 0.1),
                   q.2 = quantile(age_bir, probs = 0.2),
                   q.3 = quantile(age_bir, probs = 0.3),
                   q.4 = quantile(age_bir, probs = 0.4),
                   q.5 = quantile(age_bir, probs = 0.5))
```

You can use `summarize()` at the end (or the beginning, or middle, depending on what you're doing!) of a pipe:
```{r}
nlsy %>%
  mutate(age_bir_stand = (age_bir - mean(age_bir)) / sd(age_bir)) %>%
  filter(sex == 1) %>%
  summarize(mean_men = mean(age_bir_stand))
```

### Correlation {-}

Among the only children, find the correlation between hours of sleep on weekdays and weekends.

```{r}

```

### Proportions {-}

Create a variable that is 1 if an observation has an income between 20,000 and 30,000, and 0 otherwise. Calculate the proportion of people in the dataset who fit that criterion.

```{r}

```


### `summary()` reimagined {-}

Recreate the `summary()` function using `summarize()` (i.e., produce all the same statistics) for a variable of your choice.

```{r}

```

## Stratified summary statistics

Note the grouping information that gets printed out when you use `group_by()`:
```{r}
# grouped dataset
nlsy %>%
  mutate(income_stand = (income - mean(income))/sd(income)) %>%
  select(id, region, income_stand, race_eth, sex) %>%
  group_by(race_eth)
```

Now functions within `summarize()` will be applied within the groups:
```{r}
nlsy %>%
  mutate(income_stand = (income - mean(income))/sd(income)) %>%
  group_by(region) %>%
  summarize(mean_inc = mean(income_stand),
            sd_inc = sd(income_stand))
```

(The message is just telling us that the new dataset is no longer grouped, and that we can change that using the `.groups = ` argument.)

You can group by multiple variables:
```{r}
nlsy %>%
  group_by(region, sex) %>%
  summarize(mean_inc = mean(income),
            sd_inc = sd(income))
```

When you group by multiple variables and then summarize, the default is to "peel off" the last grouping layer and leave the rest, as the message describes. This can be confusing/annoying if you don't realize it, so you might want to use `.groups = "drop"`.

Here are three different ways to do the same thing (a common "problem" in R!):
```{r, message = FALSE}
nlsy %>%
  group_by(sex) %>%
  summarize(n = n())

nlsy %>%
  group_by(sex) %>%
  tally()

nlsy %>% 
  count(sex)
```

You can use any one of them to calculate proportions:
```{r, message = FALSE}
nlsy %>%
  group_by(sex) %>%
  summarize(n = n()) %>%
  mutate(prop = n / sum(n))
```

### By region {-}

Find the median income per region. Before doing so, make sure that you've made region into a factor variable with appropriate names so we can easily read your results.

```{r}

```

### By a new variable {-}

Calculate and compare the median income for people who sleep at least 8 hours on weekdays and those who sleep less.

```{r}

```

### Proportions by region {-}

Among the women (`sex = 2`), calculate the proportion who live in each region.

```{r}

```

## Tables with the `tableone` package

If you have never installed the `tableone` package, you'll need to do that first:
```{r, eval = FALSE}
install.packages("tableone")
```

Then load the library and create a table 1!
```{r}
library(tableone) # in general, put at the top of your script

tab1 <- CreateTableOne(
  data = nlsy,
  vars = c("eyesight", "nsibs", "race_eth",
           "sex", "region", "income", "age_bir"),
  strata = "glasses",
  factorVars = c("eyesight", "race_eth", "sex", "region")
)

tab1
```

Not only are there many possible arguments to the `CreateTableOne()` function, but also for the `print()` function when applied to a `TableOne` object. You can look up the documentation for both:
```{r}
print(tab1, catDigits = 2, contDigits = 2, test = FALSE, smd = TRUE)

help("CreateTableOne")
help("print.TableOne")
```

You can use this code to help get your table 1 in an easy-to-share format:

```{r, eval = FALSE}
table1_to_share <- tab1 %>% 
  print() %>% 
  as_tibble(rownames = "id")
write_csv(table1_to_share, "table1.csv")
```

Check out some other packages for tables as well: 
https://github.com/kaz-yos/tableone#similar-or-complementary-projects.

### Prepare a dataset {-}

Apply the following inclusion criteria to the NLSY dataset to make a analysis dataset: from region 1 or 4, with at least 2 siblings, and doesn't wear glasses.

```{r}

```

### Make a stratification variable {-}

Make a variable in that dataset that categorizes people in quartiles (split at 25th, 50th, and 75% percentiles of the new dataset) by income. Make sure the categories have descriptive names.

```{r}

```

### Make table 1 {-}

Make a table 1 for this new dataset, stratified by the new income variable. Make sure they show up in the correct order in your table. Include p-values testing across strata but only print 2 digits for them. Perform an exact test for the region comparison.

```{r}

```

### Share your table 1 {-}

Using the code above as a starting point, save your table 1 as a csv file.

```{r}

```

Open up the csv file in another program (Excel, Google Sheets, etc.) and check it out!