---
title: "Week 3 Exercises"
output:
  html_document:
  # prettydoc::html_pretty:
  #   theme: architect
  #   highlight: vignette
    toc: true
    toc_depth: 2
    number_sections: true
    pandoc_args: ["--number-offset=2"]
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  error = TRUE,
  collapse = TRUE,
  comment = "#> ",
  class.output = "output",
  class.message = "message"
)
```

Then you'll need to load the `tidyverse` package and read in the data like before. I included it in this week's materials too, so we don't have to worry about reading it in from a different directory:
```{r}
library(tidyverse)
nlsy <- read_csv("nlsy_cc.csv")
```


## Making new variables with `mutate()`

### Standardized income {-}

Using the NLSY data and `mutate()`, make a standardized (centered at the mean, and divided by the standard deviation) version of income.

```{r}
# replace the ... with your code
nlsy <- mutate(nlsy, income_stand = ...)
```

### Standardized log(income) {-}

Do the same thing, but using income on the log scale. Look at this variable using `summary()`. Can you figure out what happened? (Hint: look at your log(income) variable.)

```{r}
nlsy <- mutate(nlsy, log_income_stand = ...)
summary(nlsy$log_income_stand)
```

### `case_when()` {-}

Redo the previous question, but if you are not able to calculate log(income) for an observation, replace it with a missing value (using `case_when()`). This time, when you standardize log(income), you'll have to use `na.rm = TRUE` to remove missing values both when you take the mean and the standard deviation.

```{r}
nlsy <- mutate(nlsy,
               log_income_stand = case_when(
                 ...
               ))
```

## Factors in R using the `forcats` package

### Recode a factor {-}

Turn the eyesight variable into a factor variable. The numbers 1-5 correspond to "excellent", "very good", "good", "fair", and "poor." Make sure that categories are in an appropriate order.

```{r}
nlsy <- mutate(nlsy, eyesight_fact = ...)
```

### Combining factor levels {-}


Use two different methods to combine the worst two categories of eyesight into one category.

```{r}

```

### Relevel a factor {-}

Make a new categorical income variable with at least 3 levels (you can choose the cutoffs). Make a bar graph with this new variable where the bars are in the correct order from low to high.

```{r}

```

## Selecting variables using `select()`

### Select centered variables {-}

Create mean-centered versions of "age_bir", "nsibs", "income", and the two sleep variables. Use the same ending (e.g., "_cent") for all of them. Then make a new dataset of just the centered variables using `select()` and a helper.

```{r}
nlsy <- mutate(nlsy,
               age_bir_cent = ...,
               nsibs_cent = ...,
               ...)
nlsy_new <- select(nlsy, ...)
```

### Go back to the beginning {-}

You may have added a lot of variables to the original dataset by now. Create a dataset called `nlsy_orig` that contains only the variables we started off with, using the vector of names we originally used to name the columns and the `all_of()` helper. I'll start you off with the variable names.

```{r}
colnames_orig <- c("glasses", "eyesight", "sleep_wkdy", "sleep_wknd",
                   "id", "nsibs", "samp", "race_eth", "sex", "region", 
                   "income", "res_1980", "res_2002", "age_bir")


```

### Rename variables {-}

Look at `help(rename)`. Looking at the examples to help, rename "age_bir" to "age_1st_birth" without making a new column.

```{r}

```

## Subset your data with `filter()`

### "Or" conditions {-}

Create a dataset with all the observations that get over 7 hours of sleep on both weekends and weekdays *or* who have an income greater than/equal to 20,000 and less than/equal to 50,000.

```{r}
nlsy_or <- filter(nlsy, ...)
```

### Missing values {-}

Create a dataset that consists *only* of the missing values in `slp_cat_wkdy`. Check how many rows it has (there should be 3!).

```{r}

```

### Greater than/less than {-}

Look up the `between()` function in help. Figure out how to use this to answer the first question in this section, when choosing people whose income is between 20,000 and 50,000. Check to make sure you get the same number of rows.

```{r}

```